Embedding to Reference t-SNE Space Adresses Batch-Effects in Single-Cell Classification

### 2019-06-13 paper/tsne_embedding.tex
Corrections and edits.

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data sets,
the straightforward application of these methods often fails, and instead of
revealing underlying classes, the resulting visualizations expose
domain-specific clusters. To circumvent these batch effects, we here propose an
embedding procedure that takes a t-SNE visualization constructed on reference
data set and uses it as a scaffold for new embeddings. New, secondary data is
embedded one data-instance at the time, thus disregarding any interactions
between data items in the secondary data and implicitly mitigating batch
effects. We demonstrate the utility of this procedure on analysis of six
recently published single-cell gene expression data sets containing up to tens
of thousands of cells and thousands of genes. In these data sets, the batch
effects were particularly strong as the data comes from different institutions
and was obtained using different experimental protocols. The visualizations
constructed by the proposed approach are absent of batch effects, and the cells
from secondary data sets correctly co-cluster with cells from the primary data
sharing the same cell type.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of multi-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear local embeddings such as
t-distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing the data
from a single coherent source, these methods may encounter problems if the data
comes from different sources. Here, if executing the dimensionality reduction
procedures on a merged data, the resulting visualizations would typically
depict source-specific clusters instead of reveal clusters that are
semantically enriched, say, with same-type of data instances across different
sources. This source-specific confounding is often referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em dataset shift}~\cite{dataset_shift}. In the bioinformatics literature, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets typical suffer from source-specific
biases. Consider an example from single-cell genomics, a domain we will focus
on in this manuscript and that was --- besides current scientific challenges
--- selected also due to availability and the abundance of recently published
data. Single-cell data sets may include thousands of cells whose profiles
consist of tens of thousands of cell-specific gene expressions. Single cell
studies typically start with the analysis of cell types, where we expect that
cells of the same type would cluster in two-dimensional data
visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
t-SNE embedded data from mouse brain cells originating from visual
cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}. The figure reveals
distinct clusters but also separates the data from the two brain regions. These
two regions share the same cell types and --- contrary to the depiction in
Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
the exploration of pancreas cells in Fig.~\ref{fig:batch_effect}.b, which
renders the data from human cell atlas~\cite{baron2016} and similarly-typed
cells from diabetic patients~\cite{xin2016}. Just like with data from brain
cells, pancreas cells cluster first according to the data source, again
resulting in an uninformative visualization that primarily exposes the batch
effect.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically pre-processed and
transformed, possibly to the latent space, so that the batch effects are
removed. The recently proposed procedures for batch-effect removal include
canonical correlation analysis~\cite{cca} and mutual
nearest-neighbors~\cite{mnn,seurat}. In these works, a proof for the success of
batch effect removal is a good mixing of cells from different sources in a
t-SNE visualization. Elimination of batch effects may require aggressive data
preprocessing which may blur the boundaries between cell types. Another problem
is also the inclusion of any new data, for which the entire data analysis
pipeline has to be rerun, usually resulting in different layout and clusters
that have little resemblance to original visualization and thus require
reinterpretation.

We propose a direct solution of rendering t-SNE visualizations that addresses
batch effects. Our approach considers one of the data sets as a {\em reference}
and aims to visually classify the cells in the other, {\em secondary data set}.
We construct the t-SNE embedding using the reference data set, and then use it
as a scaffold for the embedding of data points from the secondary data.
Embedding is performed one data point at a time. Independence of each new
embedding of data instances from secondary data set causes clustering landscape
to only depend on a reference scaffold, thus removing data source-driven
variation. In other words, when including new data, the scaffold inferred from
the reference data set is kept unchanged and defines the ``gravitational
field'' to independently embedded each new data item. For example, in
Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the
scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from
hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t-SNE
visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamus cells are
dispersed across entire embedding space and their cell type correctly matches
the prevailing type in reference clusters.

At the core of the proposed solution is a mapping that implements embedding of
new data to existing t-SNE visualization. In the following, we introduce t-SNE,
then describe its recently proposed multi-scale extension, and outline our
algorithm that embeds new data into a previously trained scaffold. While a
utility of such an algorithm was already hinted to in recent
publication~\cite{art_of_using_tsne}, we here provide its practical and
theoretically-grounded implementation. Considering the abundance of recent
publications in batch effect removal, we present surprising evidence that a
computationally more direct and elegant embedding procedure solves the batch
effects problem when constructing interpretable visualizations from different
data sources.

@@

We describe an end-to-end pipeline that uses t-SNE embeddings as a scaffold for
new data samples, and enables visualisation of data from different sources
while mitigating batch effects. Our proposed approach starts with t\nobreakdash
-SNE to embedd a reference data set, with the aim of constructing a
two-dimensional visualisation to facilitate interpretation and cluster
classification. We regard the resulting two-dimensional embedding as a scaffold
to which we add samples from the new, secondary data source. Each new sample is
then placed into the reference embedding and optimized independently using the
t\nobreakdash -SNE loss function. Independent embedding of each data item from
a secondary data set disregards any interactions present in that data, and
prevents forming clusters that would be data-souce specific. Below, we start
with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, emphasizing
the notation and relevant parts upon which we base our secondary data embedding
approach (Sec.~\ref{sec:transfer}).

@@

The embedding by t-SNE is local and non-linear, tailored to visualisation of
high dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
\mathbb{R}^D$ where $N$ is the number of samples in the reference data set,
t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \in \mathbb{R}^d$
where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
close in the multi-dimensional space, their corresponding embeddings
$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
as a visualization tool, $d$ is typically set to two. The similarity between
two data points in t-SNE is defined as:

@@

\noindent where $\mathcal{D}$ is some distance measure. This is then symmetrized to

@@

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
embedding space are defined using the $t$-distribution with one degree of
freedom

@@

The data transformation by t-SNE finds an embedding $\mathbf{Y}$ that minimizes
the Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the applications of
the algorithm impractical for any reasonably-sized data. To address larger data
sets, we instead adapt a recent approach for low-rank approximation of
similarity matrix based on polynomial interpolation that reduces the time
complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
visualization of massive data sets possibly containing millions of data
points~\cite{fi_tsne}.

The t-SNE embeddings substantially depend on the value of the perplexity
parameter. The perplexity parameter can be interpreted as the number of
neighbors for which the distances in the embedding space are preserved. Small
values of perplexity result in tightly-packed clusters of points and lead to
ignoring the long-range interactions between clusters. Larger values may result
in a more globally consistent visualisations, preserving distances on a large
scale and organizing clusters in a more meaningful way. Larger values of
perplexity may often result in merging multiple small clusters, thus obscuring
local aspects of the data~\cite{art_of_using_tsne}.

@@

\noindent where $L$ is the number of mixture components. The bandwidths
$\sigma_{i,l}$ are selected in the same manner as in
Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each
$l$. In our experiments, we used a mixture of two Gaussian kernels with
perplexity values of 50 and 500. We note that a similar formulation of
multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
resulting embeddings are visually very similar to those obtained with the
approach described above (for brevity, data not shown).

@@

An algorithm to embedd new data point to a reference embedding consists of
estimating similarities between the point and the reference data and optimizing
the position of the data point in the embedding space. Unlike parametric models
such as principal component analysis or autoencoders, t-SNE does not define an
explicit mapping to embedding space, and embeddings need to be found through
loss function optimization.

For each new data point, we first estimate its distance to each of the data
instances in the reference data set. We initialize the position of the data
point in embedding space to the median embedding position of its $k$ nearest
neighbors from the reference data set. While we found the algorithm to be
robust to choices of $k$, we use $k=10$ in our experiments.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum proposed by
Jacobs to speed up the convergence~\cite{momentum}. We run gradient descent
with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
optimization converged for all cases in our experiments.  The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(N)$.

Special care must be taken to reduce the learning rate $\eta$ as the default
value in most implementations ($\eta = 200$) may cause points to ``shoot off''
from the reference embedding. This phenomenon is caused due to the embedding to
already defined t-SNE space, where the distances between data points and
corresponding gradients of the optimization function may be quite large. When
running s standard t-SNE, points are initialized and scaled to have variance
0.0001. The resulting gradients tend to be very small during the initial phase,
resulting in stable convergence. When embedding new samples, the range of the
embedding is much larger, resulting in much larger gradients, and the default
learning rate causes points to move very far from the reference embedding. In
our experiments, we found that lowering the learning rate to $\eta \sim 0.1$
produces stable solutions. This is especially important when using the
interpolation-based approximation. Our approach places a grid of interpolation
points over the embedding space, where the number of grid points is determined
by the embedding range. Clearly, if even one point ``shoots off'' far from the
embedding, the number of required grid points escalates, substantially
increasing the runtime. The reduced learning rate suppresses this issue, and
does not slow the convergence because of an adaptive learning rate scheme and
running the optimization for a sufficient number of steps.

@@

We apply the proposed transfer learning approach to t-SNE visualizations of
single-cell data. In single-cell data sets, the data includes the variety of
cells from specific tissue and characterizes the cells through the expression
of its genes. In experiments, we considered several recently published data
sets where cells were annotated with the cell type. Our aim was to construct
t-SNE visualizations where similarly-typed cells would cluster together,
despite the differences between data sources. Below, we enlist the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. In each pair, the first data set
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. In
all cases, we relied on quality control and annotations from the original
studies. To facilitate comparisons, the cell annotations were harmonized using
the cell type annotations from cell ontology~\cite{cell_ontology}. Notice that
different RNA sequencing protocols were used to estimate gene expressions. We
here report the number of cell types from each data set retained after
preprocessing. Single-cell data is sparse, typically containing less than 10\%
expressed genes per cell.}
\label{tab:datasets}
\end{table}

@@

\noindent In our experiments we use $t=0$ and $a=1$.

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/transform_brain.pdf}
\caption{Two-dimensional embedding of a reference brain cells (a) and
corresponding mapping of secondary data from hypothalamus cells (b). Notice
that the majority of hypothalamus cells were mapped to their corresponding
reference cluster. For instance, the astrocyte cells marked with red on the
right were mapped to an oval cluster of same-typed cells denoted with the same
color in the visualization on the left.} \label{fig:transform_brain}
\end{figure}

@@

There are a few tricks that we use in our procedure for t-SNE embedding of the
secondary data set that were proposed recently and enhance the original t-SNE.
An important one is a multi-scale extension that besides local ordering of the
data points takes care about global optimization of the cluster placement. For
illustration, consider visualizations with standard and multi-scale t-SNE in
Fig.~\ref{fig:multiscale}. Notice, for instance, that in multi-scale t-SNE
(Fig.~\ref{fig:multiscale}.b) the clusters with neuron cells are clumped
together, while their placement in standard t-SNE is arbitrary
(Fig.~\ref{fig:multiscale}.a).

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed to position new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this trick as well,
but only for initialization of positions of new data instances that are subject
to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
nearest neighboring-based positioning is insufficient and may yield clumped
visualizations where the optimal positioning using the t-SNE loss function is
much more dispersed and rightfully shows a more considerable variation in the
secondary data. Some data points may also fell into the neighboring space
between differently typed clusters, while after optimization they typically
converge closer to same-typed groups.

@@

The proposed method assumes that all cell types from the secondary data set are
present in the reference. The proposed method would fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as scmap~\cite{scmap} were recently
proposed to cope with such cases and identify the cells whose type is new and
not included in the reference. Our procedure does not address such cases, and
for scaling-up to a wider collection of cell types relies on emerging
availability of large collections of the reference data such as those managed
by Human Cell Atlas initiative~\cite{hca}.

@@

Almost all recent publications of single-cell studies start with a
two-dimensional visualization of the data that exposes the diversity as well as
different types of cells from the study. While any dimensionality reduction
technique can be used to render such a visualization, different variants of
t-SNE are most often used. Due to the ability to explore biological mechanisms
at the cellular level, single-cell studies are increasingly widespread, and
their publications in the past couple of years are abundant. One of the central
tasks in single-cell studies is the classification of new cells based on
finding from previous studies. Such transfer of knowledge is often difficult
due to batch effects present in data from different sources. Solving the
problem of the batch-effects together with prevailing means of presenting
single-cell data in two-dimensional visualizations motivated the research
presented in this paper.

>>>>

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data sets,
a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose
data set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that takes a t-SNE visualization constructed on the reference
data set and uses it as a scaffold for embedding new data. New, secondary data is
embedded one data-instance at the time. This prevents any interactions
between data instances in the secondary data and implicitly mitigates batch
effects. We demonstrate the utility of this approach on analysis of six
recently published single-cell gene expression data sets containing up to tens
of thousands of cells and thousands of genes. In these data sets, the batch
effects were particularly strong as the data comes from different institutions
and was obtained using different experimental protocols. The visualizations
constructed by the proposed approach are cleared of batch effects, and the cells
from secondary data sets correctly co-cluster with cells from the primary data
sharing the same cell type.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear local embeddings such as
t-distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
from a single coherent source, these methods may encounter problems if the data
comes from different sources. Here, if executing the dimensionality reduction
procedures on merged data, the resulting visualizations would often
depict source-specific clusters instead of 
grouping same classes of data which are shared across different
sources. This source-specific confounding is commonly referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em dataset shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets often suffer from these source-specific
biases. Consider an example from single-cell genomics, a domain we will focus
on in this manuscript and that was --- besides current scientific challenges
--- selected also due to availability and the abundance of recently published
data. Single-cell RNA sequencing (scRNA-seq) data sets are the result of isolating RNA molecules from individual cells,
which serves as an estimate of the expression of cell's genes.
Single-cell studies, which can exceed thousands of cells and tens of thousands of genes,
typically start with the analysis of cell types. Here, it is generally expected that
cells of the same type would cluster together in two-dimensional data
visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
t-SNE embedded data from mouse brain cells originating from visual
cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}. The figure reveals
distinct clusters but also separates the data from the two brain regions. These
two regions share the same cell types and --- contrary to the depiction in
Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
the exploration of pancreas cells in Fig.~\ref{fig:batch_effect}.b, which
renders the data from human cell atlas~\cite{baron2016} and
related cells from diabetic patients~\cite{xin2016}. Just like with data from brain
cells, pancreas cells cluster primarily by data source, again
resulting in an uninformative visualization driven by batch
effects.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems upfront. The data is first preprocessed and
transformed such that the batch effects are explicitly
removed. Recently proposed procedures for batch effect removal include
Canonical correlation analysis~\cite{cca} and Mutual
nearest-neighbors~\cite{mnn,seurat}. In these works, a proof for the success of
batch effect removal is a good mixing of cells from different sources in a
t-SNE visualization. Elimination of batch effects may require aggressive data
preprocessing which may blur the boundaries between cell types. Another problem
is also the inclusion of any new data, for which the entire analysis
pipeline has to be rerun, potentially resulting in different layout and clusters
that have little resemblance to original visualization and thus require
reinterpretation.

We propose a direct solution of rendering t-SNE visualizations that addresses
batch effects. Our approach considers one of the data sets as a {\em reference}
and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
We construct the t-SNE embedding using the reference data set, and then use it
as a scaffold for the embedding of secondary data points.
A key idea underpinning our approach is that the embedding is performed one data point at a time. 
Independence of each new
embedding of data instances from secondary data set causes the clustering landscape
to only depend on the reference scaffold, implicitly removing data source-driven
variation. In other words, when including new data, the scaffold inferred from
the reference data set is kept unchanged and defines a ``gravitational
field'', independently driving the embedding of each new instance. For example, in
Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the
scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from
hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t-SNE
visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamus cells are
dispersed across entire embedding space and their cell type correctly matches
the prevailing type in reference clusters.

The proposed solution is implemented using a mapping of
new data to an existing t-SNE visualization. While the 
utility of such an algorithm was already hinted to in recent
publication~\cite{art_of_using_tsne}, we hereby provide its practical and
theoretically-grounded implementation. Considering the abundance of recent
publications on batch effect removal, we present surprising evidence that a
computationally more direct and principled embedding procedure solves the batch
effects problem when constructing interpretable visualizations from different
data sources.

@@

We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a scaffold for
embedding new (secondary) data, enabling joint visualisation of multiple data sources
while mitigating batch effects. Our proposed approach starts with t\nobreakdash
-SNE to embed the reference data set, with the aim of constructing a
two-dimensional visualisation to facilitate interpretation and cluster
classification. Then, the placement of each new sample is
optimized independently via the
t\nobreakdash -SNE loss function. Independent treatment of each data instance from
a secondary data set disregards any interactions present in that data set, and
prevents forming clusters that would be specific to the secondary data. Below, we start
with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, emphasizing
the notation and relevant parts upon which we base our secondary data embedding
approach (Sec.~\ref{sec:transfer}).

@@

The non-linear t-SNE embedding emphasizes local similarities between data instances, 
as it is tailored to visualisation of high dimensional data sets. 
Given a multi-dimensional data set $\mathbf{X} =
\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
\mathbb{R}^D$ where $N$ is the number of samples in the reference data set,
t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \in \mathbb{R}^d$
where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
close in the multi-dimensional space, their corresponding embeddings
$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
as a visualization tool, $d$ is typically set to two. The similarity between
two data points in t-SNE is defined as:

@@

\noindent where $\mathcal{D}$ is a distance measure. This is then symmetrized to

@@

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
embedding space is defined using the $t$-distribution with one degree of
freedom

@@

The t-SNE method finds an embedding $\mathbf{Y}$ that minimizes
the Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the applications of
the algorithm impractical for any reasonably-sized data. To address larger data
sets, we adopt a recent approach for low-rank approximation of
similarity matrix based on polynomial interpolation that reduces the time
complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
visualization of massive data sets, possibly containing millions of data
points~\cite{fi_tsne}.

The resulting embeddings substantially depend on the value of the perplexity
parameter. Perplexity can be interpreted as the number of
neighbors for which the distances in the embedding space are preserved. Small
values of perplexity result in tightly-packed clusters of points and lead to
ignoring the long-range interactions between clusters. Larger values may result
in a more globally consistent visualisations, preserving distances on a large
scale and organizing clusters in a more meaningful way. Larger values of
perplexity can lead to merging of multiple small clusters, thus obscuring
local aspects of the data~\cite{art_of_using_tsne}.

@@

\noindent where $L$ is the number of mixture components. The bandwidths
$\sigma_{i,l}$ are selected in the same manner as in
Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each
$l$. In our experiments, we used a mixture of two Gaussian kernels with
perplexity values of 50 and 500. We note that a similar formulation of
multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
resulting embeddings are visually very similar to those obtained with the
approach described above (not shown for brevity).

@@

An algorithm to embed a new data point to a reference embedding consists of
estimating similarities between the point and the reference data and optimizing
the position of the data point in the embedding space. Unlike parametric models
such as principal component analysis or autoencoders, t-SNE does not define an
explicit mapping to the embedding space, and embeddings need to be found through
loss function optimization.

The position of a new data
point in embedding space is initialized 
to the median reference embedding position of its $k$ nearest 
neighbors. While we found the algorithm to be
robust to choices of $k$, we use $k=10$ in our experiments.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum 
to speed up the convergence, as proposed by Jacobs~\cite{momentum}. We run gradient descent
with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
optimization converged for all cases in our experiments.  The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(N)$.

Special care must be taken to reduce the learning rate $\eta$ as the default
value in most implementations ($\eta = 200$) may cause points to ``shoot off''
from the reference embedding. This phenomenon is caused due to the embedding to
a previously defined t-SNE space, where the distances between data points and
corresponding gradients of the optimization function may be quite large. When
running s standard t-SNE, points are initialized and scaled to have variance
0.0001. The resulting gradients tend to be very small during the initial phase,
resulting in stable convergence. When embedding new instances, the range of the
embedding is much larger, resulting in much larger gradients, and the default
learning rate causes points to move very far from the reference embedding. In
our experiments, we found that decreasing the learning rate to $\eta \sim 0.1$
produces stable solutions. This is especially important when using the
interpolation-based approximation. Our approach places a grid of interpolation
points over the embedding space, where the number of grid points is determined
by the embedding range. Clearly, if even one point ``shoots off'' far from the
embedding, the number of required grid points escalates, substantially
increasing the runtime. The reduced learning rate suppresses this issue, and
does not slow the convergence because of an adaptive learning rate scheme, provided
the optimization is run for a sufficient number of steps.

@@

We examine the utility of the proposed approach with t-SNE visualizations of
single-cell data. Data in this realm include a variety of
cells from specific tissues, and characterize them through gene expression. 
We considered several recently published data
sets where cells were annotated with the cell type. Our aim was to construct
t-SNE visualizations where cells of similar types would cluster together,
despite systematic differences between data sources. Below, we enlist the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. In each pair, the first data set
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. In
all cases, we relied on quality control and annotations from the original
studies. To facilitate comparisons, the cell annotations were harmonized using
the cell type annotations from Cell ontology~\cite{cell_ontology}. Notice that
different RNA sequencing protocols were used to estimate gene expressions. We
here report the number of cell types from each data set retained after
preprocessing. Single-cell data is sparse, typically containing less than 10\%
expressed genes per cell.}
\label{tab:datasets}
\end{table}

@@

\noindent In our experiments, we use $t=0$ and proceed with a subset of
$\hat{G}$ genes.

@@

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/transform_brain.pdf}
\caption{Two-dimensional embedding of a reference brain cells (a) and
corresponding mapping of secondary data from hypothalamus cells (b). Notice
that the majority of hypothalamus cells were mapped to their corresponding
reference cluster. For instance, the astrocyte cells marked with red on the
right were mapped to an oval cluster of the same type, with cells denoted using the same
color in the visualization on the left.} \label{fig:transform_brain}
\end{figure}

@@

There are additional modifications that we use in the embedding of the
secondary data set that were proposed recently and enhance the original t-SNE.
An important one is a multi-scale extension that, besides local ordering of the
data points, globally optimizes the cluster placement. For
illustration, consider visualizations with standard and multi-scale t-SNE in
Fig.~\ref{fig:multiscale}. Notice, for instance, that in multi-scale t-SNE
(Fig.~\ref{fig:multiscale}.b) the clusters with neuron cells are clumped
together, while their placement in standard t-SNE is arbitrary
(Fig.~\ref{fig:multiscale}.a).

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed to position new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for initialization of positions of new data instances that are subject
to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
nearest neighboring-based positioning is insufficient and may yield clumped
visualizations where the optimal positioning using the t-SNE loss function is
much more dispersed and rightfully shows a more considerable variation in the
secondary data. Some data points may also fell into the neighboring space
between differently typed clusters, while after optimization they typically
converge closer to groups of the same type.

@@

The proposed method assumes that all cell types from the secondary data set are
present in the reference. The proposed method would fail to reveal novel cell
types in the secondary data set, possibly positioning them arbitrarily close to
unrelated clusters. Procedures such as Scmap were recently
proposed to cope with such cases and identify the cells whose type is new and
not included in the reference~\cite{scmap}. Our procedure does not address such cases, and
for scaling-up to a wider collection of cell types relies on emerging
availability of large collections of the reference data such as those managed
by Human Cell Atlas initiative~\cite{hca}.

@@

A large majority of recent publications on single-cell studies start with a
two-dimensional visualization of the data that exposes the diversity as well as
different types of cells in the study. While any dimensionality reduction
technique can be used to render such a visualization, different variants of
t-SNE are most often used. Due to the ability to explore biological mechanisms
at the individual cell level, single-cell studies are increasingly widespread, and
their publications in the past couple of years are abundant. One of the central
tasks in single-cell studies is the classification of new cells based on
finding from previous studies. Such transfer of knowledge is often difficult
due to batch effects present in data from different sources. Solving the
problem of the batch-effects together with prevailing means of presenting
single-cell data in two-dimensional visualizations motivated the research
presented in this paper.
